---
title: "Captsone3_EDA"
output: 
 github_document:
 pandoc_args: ["--wrap=none"]
always_allow_html: true
date: "2025-04-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Set CRAN mirror first (this fixes the mirror error)
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Install required packages if not already installed
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("corrplot", quietly = TRUE)) install.packages("corrplot")

# Load basic libraries that should always work
library(tidyverse)
library(knitr)
```

# Exploratory Data Analysis of Wine Quality

## 1. Business Understanding

### What Makes a Good Quality Wine?

According to wine experts and research in oenology (the science of wine-making), several factors contribute to wine quality:

- **Chemical Composition**: Acidity levels, alcohol content, sugar levels, and presence of various compounds all impact taste and aroma
- **Balance**: The harmony between sweetness, acidity, tannin, and alcohol
- **Complexity**: Wines with multiple flavor notes are often rated higher
- **Region and Climate**: Terroir significantly impacts grape quality 
- **Aging Process**: Proper aging can develop desirable flavors and aromas

For this analysis, we'll focus on the measurable chemical properties provided in the dataset to determine which factors correlate most strongly with wine quality ratings.

## 2. Data Understanding

The UCI Wine Quality Dataset contains physicochemical properties of red wines and their quality ratings. Let's load and examine this dataset:

```{r load_data}
# Download the dataset directly rather than looking for it locally
wine_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv"
wine_data <- read.csv(wine_url, sep = ";")

# Examine the structure of the dataset
str(wine_data)

# Display the first few rows
head(wine_data)

# Summary statistics
summary(wine_data)

# Check for missing values
cat("Missing values:", sum(is.na(wine_data)))

# Describe data types
cat("\nData Types of Features:\n")
cat("1. fixed acidity: Ratio (g/L)\n")
cat("2. volatile acidity: Ratio (g/L)\n")
cat("3. citric acid: Ratio (g/L)\n")
cat("4. residual sugar: Ratio (g/L)\n")
cat("5. chlorides: Ratio (g/L)\n")
cat("6. free sulfur dioxide: Ratio (mg/L)\n")
cat("7. total sulfur dioxide: Ratio (mg/L)\n")
cat("8. density: Ratio (g/cm³)\n")
cat("9. pH: Ratio (scale from 0-14)\n")
cat("10. sulphates: Ratio (g/L)\n")
cat("11. alcohol: Ratio (% by volume)\n")
cat("12. quality: Ordinal (score between 0-10)\n")
```

### Dataset Description:

This dataset contains samples of red wine with various physicochemical properties and quality ratings. The dataset comes from the UCI Machine Learning Repository and was created to model wine preferences based on these properties.

## 3. Data Preparation

Let's prepare our data for analysis by checking for duplicates and creating some additional variables:

```{r data_prep}
# Check for duplicated rows
cat("Number of duplicated rows:", sum(duplicated(wine_data)))

# Convert quality to a factor for categorical analysis
wine_data$quality_factor <- factor(wine_data$quality, 
                                  ordered = TRUE,
                                  levels = unique(sort(wine_data$quality)))

# Create a binary variable for "good quality" wine (quality >= 7)
wine_data$high_quality <- ifelse(wine_data$quality >= 7, "high", "low")
wine_data$high_quality <- factor(wine_data$high_quality)

# Show class distribution
table(wine_data$quality)
table(wine_data$high_quality)
```

## 4. Modeling + Evaluating (Exploratory Data Analysis)

### 4.1 Quality Distribution

Let's first look at the distribution of wine quality ratings:

```{r quality_dist}
ggplot(wine_data, aes(x = quality_factor)) +
  geom_bar(fill = "darkred") +
  labs(title = "Distribution of Wine Quality",
       x = "Quality Score",
       y = "Count") +
  theme_minimal()
```

### 4.2 Exploring Chemical Properties

Let's see how chemical properties are distributed:

```{r chemical_dist}
# Create histograms for key variables
wine_data %>%
  select(fixed.acidity, volatile.acidity, citric.acid, alcohol) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(fill = "steelblue", bins = 30) +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Distribution of Key Chemical Properties") +
  theme_minimal()
```

### 4.3 Correlation Analysis

Let's examine how different chemical properties correlate with each other and with wine quality:

```{r correlation, message=FALSE, warning=FALSE}
# Load the corrplot library if available
if(requireNamespace("corrplot", quietly = TRUE)) {
  library(corrplot)
  
  # Correlation matrix of chemical properties
  correlation_matrix <- cor(wine_data[, 1:12])
  corrplot(correlation_matrix, method = "circle", type = "upper", 
           tl.col = "black", tl.srt = 45,
           title = "Correlation Matrix of Wine Properties")
}

# Correlation with quality specifically
quality_cor <- cor(wine_data[, 1:11], wine_data[, "quality", drop = FALSE])
quality_cor_df <- data.frame(
  Variable = colnames(wine_data)[1:11],
  Correlation = as.numeric(quality_cor)
)

# Sort by absolute correlation
quality_cor_df <- quality_cor_df[order(abs(quality_cor_df$Correlation), decreasing = TRUE), ]

# Plot correlations with quality
ggplot(quality_cor_df, aes(x = reorder(Variable, abs(Correlation)), y = Correlation)) +
  geom_bar(stat = "identity", aes(fill = Correlation > 0)) +
  scale_fill_manual(values = c("firebrick", "steelblue"), 
                    labels = c("Negative", "Positive")) +
  coord_flip() +
  labs(title = "Correlation of Variables with Wine Quality",
       x = "Chemical Properties",
       y = "Correlation Coefficient",
       fill = "Direction") +
  theme_minimal()
```

### 4.4 Property Comparisons Across Quality Levels

Let's compare key chemical properties across different quality levels:

```{r boxplots}
# Get top 4 correlated variables
top_vars <- head(quality_cor_df$Variable, 4)

# Create and print boxplots for top variables
for (var_name in top_vars) {
  p <- ggplot(wine_data, aes_string(x = "quality_factor", y = var_name)) +
    geom_boxplot(fill = "lightblue") +
    labs(title = paste(var_name, "Across Wine Quality Levels"),
         x = "Wine Quality",
         y = var_name) +
    theme_minimal()
  print(p)
}
```

### 4.5 Statistical Tests

Let's perform some basic statistical tests to determine if there are significant differences in chemical properties between high-quality and low-quality wines:

```{r stat_tests}
# Test the relationship between alcohol content and high quality (t-test)
alcohol_ttest <- t.test(wine_data$alcohol ~ wine_data$high_quality)
print(alcohol_ttest)

# Test for volatile acidity (another highly correlated variable)
va_ttest <- t.test(wine_data$volatile.acidity ~ wine_data$high_quality)
print(va_ttest)
```

### 4.6 Simple Predictive Model

```{r simple_model}
# Simple linear regression model
simple_model <- lm(wine_data$quality ~ wine_data$alcohol + wine_data$volatile.acidity + 
                    wine_data$sulphates + wine_data$citric.acid)
model_summary <- summary(simple_model)
print(model_summary)

# Check model assumptions
par(mfrow = c(2, 2))
plot(simple_model)
par(mfrow = c(1, 1))
```

## 5. Deployment (Findings and Insights)

Based on our analysis, here are the key findings about factors affecting wine quality:

### Key Factors Affecting Wine Quality

```{r key_factors_table}
# Create a table of coefficients from our linear model
coefficients_table <- as.data.frame(model_summary$coefficients)
kable(coefficients_table, caption = "Influence of Chemical Properties on Wine Quality")
```

### Summary of Findings:

1. **Alcohol Content**: Our analysis shows that higher alcohol content is strongly associated with higher quality ratings. The t-test confirms a statistically significant difference in alcohol content between high and low-quality wines (p < 0.001).

2. **Volatile Acidity**: Lower volatile acidity (which creates vinegar-like taste) strongly correlates with higher quality wines. This is one of the most important negative factors in determining wine quality.

3. **Sulphates**: Higher sulphate levels are associated with higher quality. Sulphates serve as antimicrobials and antioxidants in wine, which helps preserve flavor and freshness.

4. **Citric Acid**: Moderate amounts of citric acid positively correlate with quality, likely due to its role in providing freshness to the wine's flavor profile.

5. **Sugar-Acidity Balance**: The interaction between residual sugar and acidity appears important for overall balance, which affects quality perception.

### Actionable Recommendations for Wine Producers:

1. **Control Volatile Acidity**: Implement better control measures during fermentation to keep volatile acidity levels low, as this has a strong negative impact on quality perception.

2. **Alcohol Management**: Consider fermentation techniques that can produce higher alcohol content where appropriate, as this is positively associated with quality ratings.

3. **Sulphate Addition**: Optimize sulphate additions, as higher levels (within legal limits) can positively impact quality through better preservation and flavor development.

4. **Acidity Balance**: Pay careful attention to the balance between different acid types (fixed, volatile, and citric) as this affects the overall taste profile significantly.

5. **Quality Testing**: Implement regular testing of these key chemical properties as quality control checkpoints during the production process.

These findings provide a data-driven approach to wine production that could potentially improve quality ratings and consumer satisfaction.
